From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-02 Pi+/paper/finite.tex

\[
  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
  \resizebox{0.5}{!}{
    \begin{center}
      \begin{tikzpicture}
        \pic[local bounding box=my braid,braid/.cd,
          number of strands = 2,
          thick]
        {braid={ s_1, s_1}};
      \end{tikzpicture}
    \end{center}
     &
    \resizebox{2.0}{!}{\(\xlongrightarrow[]{\cancel}\)}
     &
    \begin{center}
      \begin{tikzpicture}
        \pic[local bounding box=my braid,braid/.cd,
          number of strands = 2,
          thick]
        {braid={1, 1}};
      \end{tikzpicture}
    \end{center}
       }
  \end{tabular}
\]

@@

\jk{maybe change that?}
Group persentation, as given in the ~\label{def:presentation}, requires the relation to be on the set of $A + A$, where the right copy corresponded to the set of formal inverses of the generators. However, in our case, we can observe that that the constructor $\cancel$ specifies that the inverse of each element is again the same element. Thus, the type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.

@@

The approach to solving the word problem for $\Sn$ we use is to turn the relations $\cox*$ on
$\List[\Fin[n]]$ to a rewriting system $(\List[\Fin[n]],\cox*)$. A normal form of an element can then be computed by following the rewrite rules, and two words can be compared for $\cox*$-equality by comparing their normal forms.

A well-behaved rewriting system has to be strongly normalising and confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$. Thus, by directing the relation from left to right, we would get the termination property out of the box.

@@

Because of the counter-example discussed, the relations have to be changed. In this section, we will formally define a rewriting system partially based on the Coxeter relation, prove that it has the desired properties of confluence and strong normalisation.

Fortunately, the change from the rewriting system described before will be small. Thus, we will not lose the connection to the theory of group presentations, where Coxeter relations are a standard notion~\todo{citation} -- to make the result broadly applicable, even though we use the reduction system for computation, we prove that the \jk{relation it gives} is indeed equivalent to the conguence closure of the standard Coxeter relation $\cox$.

The new reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$. We fix the problem of the non-converging critical pair discussed previously by changing $\braid$ to be a slightly more general relation $\longbraid$.

\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}

@@

The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\). Since we think of the number $m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left, shifting all the elements in between one place right. Expressed in terms of the braiding diagram, for
$n = 0$ and $k = 4$, it has the following form:

@@

Constructors $\longcancel$ and $\longswap$ correspond directly to the approproiate constructors of $\cox$ and can be visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 4$, it allows for the reduction
$[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualized on the braid diagram, it looks as follows:

@@

Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$.

Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity.

@@

The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and, as we mentioned, have little connection to the broader scope of computational group theory. However, we can prove that the closures of the two forms of the relations are equivalent, and work up to this equivalence!

\begin{proposition}~\label{prop:coxlongcox}
  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$ such that $w \longcox* u$ and $v \longcox* u$.
\end{proposition}
\jk{does the construction above has a name? It is analogous to the one we do in lambda calculus, when taking $\equiv_{\beta}$. We have to write something about this proof using the Church-Rosser idea.}

By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have
the same normal form. We state a few important properties that $\normf$ satisfies.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
  \end{enumerate}
\end{proposition}

@@

\begin{proposition}
  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
\end{proposition}

@@

For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as the number of elements smaller than it appearing after in the permutation.
\todo{typify}
\begin{definition}
    Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
    $|\{j < i : \sigma(j) > \sigma(i)\}|$.
\end{definition}

@@

The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of $\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it is last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 1, 2, 0, 1)$.

@@

\todo{So now, finally, give an explict example of a normal form, Lehmer immersion}.

>>>>

\[
  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
    \begin{center}
      \begin{tikzpicture}
        \pic[local bounding box=my braid,braid/.cd,
          number of strands = 2,
          thick]
        {braid={ s_1, s_1}};
      \end{tikzpicture}
    \end{center}
     &
    \(\xlongrightarrow[]{\cancel}\)
     &
    \begin{center}
      \begin{tikzpicture}
        \pic[local bounding box=my braid,braid/.cd,
          number of strands = 2,
          thick]
        {braid={1, 1}};
      \end{tikzpicture}
    \end{center}
  \end{tabular}
\]

@@

The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,

@@

A group presentation, as defined in~\label{def:presentation}, requires the relation to be on the set of words $A + A$,
where the right copy corresponds to the set of formal inverses of the generators. We will resolve this mismatch later
in~\cref{prop:lift-codiag}.

\todo{check: However, in our case, the constructor $\cancel$ specifies that the inverse of each element is again the
  same element.}

To solve the word problem for $\Sn$, we turn the relations into a rewriting system $(\List[\Fin[n]],\cox*)$. A normal
form of an element can then be computed by following the rewrite rules, and two words can be compared for
$\cox*$-equality by comparing their normal forms. A well-behaved rewriting system has to be strongly normalising and
confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations
$\cox*$ are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$.
Thus, by directing the relation from left to right, we would get the termination property out of the box.

@@

Because of the counter-example discussed, the relations have to be changed. In this section, we will formally define a
rewriting system partially based on the Coxeter relations, and prove that it has the desired properties of confluence
and strong normalisation.

The Coxeter relations are a standard notion in the theory of group presentation~\todo{citation} -- even though we change
the relations a bit, we will not lose the connection. In fact, we will prove that the new relations are equivalent, in a
technical sense, to the standard Coxeter relations $\cox*$.

The new reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$.
We fix the problem of the non-converging critical pairs discussed previously by changing $\braid$ to be a slightly more
general relation $\longbraid$. \review{Further, we also inline the congruence closure in $\longcox$, allowing arbitrary
  reductions inside the list.}

\todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}

@@

The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\). Since we think of the number
$m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to
produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ $k$ places left, shifting all
the elements in between one place right. Expressed in terms of the braiding diagram, for $n = 0$ and $k = 4$, it has the
following form:

@@

Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and can be
visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 4$, it allows for
the reduction $[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualised on the braid diagram, it looks as follows:

@@

Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$. Also, as before, the left-hand
sides of the relation are strictly larger than the right-hand sides, in terms of the lexicographic order on
$\List[\Fin[n]]$.

We define the relation $\longcox*$ to be the closure of $\longcox$, under reflexivity and (right-step extended)
transitivity.

@@

The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction. However,
we can use the following observation to be able to use the more standard $\cox*$ relation.~\todo{Check Huet's paper for
  this property.}

\begin{proposition}~\label{prop:coxlongcox}
  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$
  such that $w \longcox* u$ and $v \longcox* u$.
\end{proposition}

By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are defined to be those terms that
have the same normal form. We state a few important properties that $\normf$ satisfies.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
    \item $\normf$ splits into a section-retraction pair, that is, we have
          ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$
          and $r \comp s \htpy \idfunc_{\Sn[n]}$.
    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
  \end{enumerate}
\end{proposition}

\todo{We set out to solve the word problem for $\Sn$ \ldots}
\todo{Revise this section by introducing $\Sn$ here.}

@@

\begin{proposition}~\label{prop:lift-codiag}
  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
\end{proposition}

@@

For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
the number of smaller elements appearing after it in the permutation.

\todo{typify}
\begin{definition}
  Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
  $|\Set{j < i | \sigma(j) > \sigma(i)}|$.
\end{definition}

@@

The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of
$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$
is the 5-tuple $l = (0, 1, 2, 0, 1)$.

@@

\todo{So now, finally, give an explicit example of a normal form, Lehmer immersion}.
